---
title: "data cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
#The message = false code tells r not to display the message it outputs in the r markdown doc
library(dplyr)
library(tidyr)
```


##Data Sources -- This hads headers

Read in the data from the [KNB](https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1)

```{r}
catch_original = read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

#can put a link in for a download file into the read.csv code
glimpse(catch_original)
```



##Clean and Reshape data


- remove unnecessary columns

```{r}
catch_data = catch_original %>% #selecting the columns that we want to keep
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
  select(-All, -notesRegCode) #this code will instead remove the columns you don't want

glimpse(catch_data)

```

- check column types

```{r}
catch_clean = catch_data %>% 
  mutate(Chinook = as.integer(Chinook))#for some reason chinook read in as a character so this changes it to a number
glimpse(catch_clean)
```
the above gave a warning that the conversion of chinook returned a NA for some things. Need to find where that happens so you can fix it 

```{r}
i <- which(is.na(catch_clean$Chinook))
i #this tells you that the row that produces an NA value is row 401

catch_original[i, ] #this shows you the data where NA pops up and it is b/c there is an I in year 1955 instead of a 1
```

```{r}
catch_clean = catch_data %>% 
  mutate(Chinook = if_else (Chinook == "I", "1", Chinook)) %>%  #first replace the I with the character 1, then need to change the 1 to integer
  mutate(Chinook = as.integer(Chinook)) #could have actual used as.numeric too, but we stuck with the integer to be consistent with the rest of the values. difference is that as.numeric allows for decimals
 

```


- reshape data
We know that we are going to want the data in long format so this set will convert it from wide to long format

```{r}
catch_long = catch_clean %>% #this grabs the cleaned dataset with the updated variables and removed All column
  pivot_longer(cols = -c(Region, Year), names_to = "species", values_to = "catch" )#first arg is the data which we piped in, second is the columns you want to pivot over. We used the -c code to say that you don't want the region or year to move to different columns in a long format
#names_to is what you are going to turn the species column names into. Values_to is what you are naming the column where the values are inputted

head(catch_long) #head may be a better way to look at it since it has the actual format with the long format, or you could just click on it on the right

```
```{r}
#moving back to a wide format

catch_wide = catch_long %>% 
  pivot_wider(names_from = Region, values_from = catch)#this actual gives where each species was found and the amount per region

head(catch_wide)
```


```{r}
#can also rename columns, in this instance want to include the units since it is in thousands

catch_long <- catch_long %>% 
  rename(catch_thousands = catch)


  mutate(catch = catch_thousands * 1000)
```

- group

```{r}
mean_region = catch_long %>% 
  group_by(Region, Year) %>% 
  summarise(catch_mean = mean(catch), num_obs = n())

head(mean_region)
```


##Join to Region Definitions

Using a left join--- want to always make sure that you have the same amount of rows in the left table as the new table. 
The number of columns will be the number of columns in the left, right, minus what you're joining


```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)
```

```{r}
catch_joined = left_join(catch_long, region_defs) #this gives an error because the names for the columns are not the same between the two

#can help this by using "by" in the code to specify what the column name in the left is and how to joins to the right

catch_joined = left_join(catch_long, region_defs, by = c("Region" = "code"))

#another option to join is by using the function merge, but the left or right join is more specific as to how it is joined

head(catch_joined)
```








